Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study

### 2015-02-22 chemotaxis/paper/draft1.tex
updated chemotaxis paper

@@

\subsection{Manifold learning algorithm: diffusion maps}

Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$. 
%
Assume we are given $m$ observations $z(1), \dots, z(m)$, which lie on a $d$-dimensional, nonlinear manifold $\mathcal{M}_d$, with $d \ll n$. 
%
Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
%
We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
We typically take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ provide the embedding coordinates for the data, such that
$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
%
The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g., the relative importance of $\phi_1$ versus $\phi_2$ can be measured by comparing $\mu_1$ and $\mu_2$.
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the leading eigenvectors to adequately describe the data.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.

In the limit of infinite data ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplacian operator on the manifold $\mathcal{M}_d$ \cite{...}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ approximate the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$.
%
The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues we compute in diffusion maps by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = e^{-\tilde{\mu}_k t}.
\end{equation}

\subsection{Constructing appropriate observers}

\subsection{Defining a meaningful metric}

\section{Identifying the Informative Eigenvectors}

In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the top eigenvectors provide us with a parameterization of the underlying manifold.
%
However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set. 
%
We will illustrate this issue with a simple two-dimensional strip where we know the analytical form of the eigenfunctions and can explicitly highlight the issue at hand.

Consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
%
The eigenvalues of the Laplacian with Neumann boundary conditions are given by
\begin{equation}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
\end{equation}
where $x$ and $y$ denote the two coordinates of the rectangular domain. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold. 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; the eigenvalues $\tilde{\mu}_k$ are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}). 
%
In fact, if $L_x > 2 L_y$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction. 
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
%
Looking at the diffusion maps eigenvalues which correspond to modes which are {\em not} higher harmonics will then provide us with a measure of the relative lengths of the data set along the principal directions.

@@

\subsection{Algorithm: Local Linear Regression}

@@

%\subsection{Stochastic particles}

@@

>>>>

\item For the example of a square/strip, we know everything analytically, and identifying the ``true'' principal components is a problem/not obvious, and depends both on geometry and density

\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of identifying the ``true'' components due to geometry.

@@

\subsection{Manifold learning and the Laplace operator}

We consider a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$, and we are interested in uncovering a parameterization of this manifold. 
%
Many algorithms have been proposed for such a task.
%
Historically, principal component analysis (PCA) \cite{shlens2005tutorial} has been used to parmeterize linear manifolds (planes and hyperplanes), where the principal components parameterize the underlying hyperplane.
%
However, in the case when the manifold is curved, such linear techniques can fail to capture the underlying structure.

In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator on the manifold with Neumann boundary conditions \cite{....} can provide coordinates which parameterize the manifold. 
%
To provide intuition as to why this is the case, consider a two-dimensional strip with edge lengths $L_x$ and $L_y$. 
%
The eigenvalues of the Laplacian with Neumann boundary conditions are given by
\begin{equation}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
\end{equation}
where $x$ and $y$ denote the two coordinates of the rectangular domain. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold. 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.

Although these eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often yield appropriate/good coordinates to parameterize the underlying manifold.

@@

\subsection{Discrete approximation of the Laplacian: diffusion maps}

In most applications, we are not given a description of the continuous manifold. 
%
Instead, we are often given data {\em sampled} from the underlying manifold $\mathcal{M}_d$, and we would like to uncover a parameterization of the data.
%
We will do this by constructing a matrix which approximates the Laplace-Beltrami operator with Neumann boundary conditions. 
%
The eigenvectors of this matrix then approximate the eigenfunctions of this operator.

Assume we are given $m$ observations $z(1), \dots, z(m) \in \mathbb{R}^n$, which lie on a $d$-dimensional, nonlinear manifold $\mathcal{M}_d$, with $d \ll n$. 
%
We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
We typically take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{...}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$, such that
$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
%
The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues of $A$, $\mu_0, \mu_1, \dots, \mu_{m-1}$, by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = e^{-\tilde{\mu}_k t}.
\end{equation}
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the leading eigenvectors to adequately describe the data.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.

However, the data need not be sampeld uniformly from the underlying manifold.
%
If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $U$, then the matrix $I-A$ approximates the Fokker-Planck operator with potential $2U$, e.g., $(I-A) \phi \rightarrow \nabla^2 \phi - 2\nabla\phi \cdot \nabla U$.

TODO: add two sets of figures: one with data uniformly sampled, one with data with Gaussian distribiution

Common practice is to use these tools for nonlinear data where these things are coupled. Show swiss roll and torus.

\subsection{Identifying the Informative Eigenvectors using Local Linear Regression}

In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the top eigenvectors provide us with a parameterization of the underlying manifold.
%
However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set. 
%
Again, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
%
The eigenvalues of the Laplacian with Neumann boundary conditions are given by
\begin{equation}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
\end{equation}
where $x$ and $y$ denote the two coordinates of the rectangular domain. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold. 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; the eigenvalues $\tilde{\mu}_k$ are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}). 
%
In fact, if $L_x > 2 L_y$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction. 
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
%
Looking at the diffusion maps eigenvalues which correspond to modes which are {\em not} higher harmonics will then provide us with a measure of the relative lengths of the data set along the principal directions.

@@

\subsection{Problem description}

@@

\subsection{Observers and Metrics}

\subsubsection{Constructing appropriate observers}

\subsubsection{Defining a meaningful metric}

Note: histograms are not essential for computing EMD, however, they make it computationally feasible. Furthermore, histograms are good because they are robust to noise (cite PNAS).

\subsection{Results}
